Figures

### 2025-01-27 Figures/PaperFiguresextended.Rmd
Added comments

dir_out <- "/home/common/data/output/projects/ENDO/E044/A033/"
dir.create(dir_out, recursive = TRUE)
setwd(dir_out)
```

In script
Replaced ProlifPeriSecFinal with MenstrualCyclePhase_main

```{r save downsampled data}
data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033.rds")

# Downsample the number of cells per identity class
Idents(data) <- data$sample
data_ds <- subset(x = data, downsample = 200)
saveRDS(data_ds, "/home/common/data/output/projects/ENDO/E044/A033/dataA003A033_downsampled.rds")

#Some plotting
DimPlot(data, group.by = "AnnotationMain", raster = FALSE)
DimPlot(data, group.by = "Symphony_Refined_A029", raster = FALSE)
DimPlot(data, group.by = "Symphony_Refined_A029_United", raster = FALSE)

#Save object subsetted for ScaiNet samples
ScaiNETsamples <- read.csv("/home/common/data/output/projects/ENDO/E041/A001/summary_plots_cv_splits/metadata_prolif_eval_fold_all.csv")$names
Idents(data) <- data$sample
subset <- subset(data, idents = ScaiNETsamples)
rm(data)
gc()
meta <- subset@meta.data
saveRDS(subset, "/home/common/data/output/projects/ENDO/E044/A033/dataA003A033_ScaiNETsamples.rds")
rm(subset)
gc()

meta$AnnotationMain

@@

Remove? TODO
```{r order of samples by menstrual cycle phase and pseudotime}
meta_final <- readRDS("/home/common/data/output/projects/ENDO/E044/A074/dataA003A033_metadata_ptime_snn_Final.rds")

MenstrualCyclePhase_sampleOrder <- meta_final %>%
  arrange(MenstrualCyclePhase,epithelial.pseudotime) %>%
  distinct(sample,epithelial.pseudotime, MenstrualCyclePhase,CycleDay) %>%
  pull(sample)
saveRDS(MenstrualCyclePhase_sampleOrder, paste0(dir_out,"sampleorder_MenstrualCyclePhase_greenCirclemedian.RDS"))
```

Add additional data to the Seurat metadata and reformat
```{r plotting of cell types per sample}
#Set directory links and create directories
dir_data <- "L:/GitHubDesktop/endometriosis_endometrium_scRNA_atlas/_Data/"
dir_out <- "L:/GitHubDesktop/endometriosis_endometrium_scRNA_atlas/Figures_out/"
if (!dir.exists(dir_out)) {
  dir.create(dir_out)
}

#Load data
#A003A033 <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033_metadata_ptime_snn_Final.rds")
meta <- readRDS(paste0(dir_data, "EndoAtlas_GEO_meta_complete.rds"))

##Modify menstrual cycle information
# Create column MenstrualCyclePhase_main based on MenstrualCyclePhase by merging the secretory sub-phases from 
meta$MenstrualCyclePhase_main <- as.character(meta$MenstrualCyclePhase)
meta$MenstrualCyclePhase_main[meta$MenstrualCyclePhase_main %in% c("secretory_early", "secretory_mid", "secretory_late")] <- "secretory"
#Factor samples by MenstrualCyclePhase_main for plotting
meta$MenstrualCyclePhase_main <- factor(meta$MenstrualCyclePhase_main, levels = c("proliferative","periovulatory","secretory"))

##Add additional metadata from manuscript Supplementary Table 1
# Read the tab-delimited text file
SupplementaryTable1 <- read.delim(paste0(dir_data, "SupplementaryTable1.txt"), header = TRUE, stringsAsFactors = FALSE)
# Perform a left join of the `meta` dataframe with `SupplementaryTable1`
meta <- dplyr::left_join(
  meta,
  SupplementaryTable1 %>% 
    dplyr::select(
      Patient.sample.ID,
      epithelial.pseudotime,
      sequencing.batch,
      Minor.exclusion.criteria.logical,
      Menstrual.cycle.phase.marker.analysis,
      Cell.type.frequency.analysis..Figure.2e.f.,
      DEG.analysis..Figure.3a.b.,
      Ligand.receptor.analysis..Figure.3c.h.
    ),
  by = c("sample" = "Patient.sample.ID")
)

#Double-check, TODO
table(meta$sample, meta$Minor.exclusion.criteria.logical)
prolifsamples <- meta%>%
  dplyr::filter(MenstrualCyclePhase_main == "proliferative") %>%
  select(sample) %>%
  distinct() %>%
  arrange(sample) %>%
  pull(sample)
DEGsamples <- unique(read.csv("/home/common/data/output/projects/ENDO/E044/A030/major_celltypes_Tan/celltype_freq_sample.csv")$Var2)
DEGsamples <- DEGsamples[order(DEGsamples)]
setdiff(prolifsamples,DEGsamples) #MinorExclusion samples overlap with samples excluded in DEG analysis, sample87 is late secretory

##Determine order of samples by MenstrualCyclePhase and epithelial.pseudotime
MenstrualCyclePhase_sampleOrder <- meta%>%
  select(sample, MenstrualCyclePhase,epithelial.pseudotime) %>%
  distinct() %>%
  arrange(MenstrualCyclePhase, epithelial.pseudotime) %>%
  mutate(MenstrualCyclePhase_sampleOrder = row_number()) %>%
  ungroup()
#Factor samples by MenstrualCyclePhase and epithelial.pseudotime for plotting
meta$sample <- factor(meta$sample, levels = c(MenstrualCyclePhase_sampleOrder$sample))

@@

Supplementary Data Figure 4 a
```{r plotting of cell types per sample}
#########################################################################
# bar plot of the number of AnnotationRefined cells in each sample
df <- meta%>%
  filter(!is.na(AnnotationMain)) %>%
  group_by(sample, AnnotationMain, AnnotationRefined, EndometriosisGrade, MenstrualCyclePhase, epithelial.pseudotime, EndometriosisStatus) %>%
  summarise(row_count = n()) %>%
  ungroup()

#Bar plot
#####use this for paper supplement
#Bar plot split by main5 
#MenstrualCyclePhase_sampleOrder <- readRDS(paste0(dir_out,"sampleorder_MenstrualCyclePhase_greenCirclemedian.RDS"))
#df$sample <- factor(df$sample, levels = MenstrualCyclePhase_sampleOrder)
P63 = createPalette(63, c("#ff0000", "#00ff00", "#0000ff")) 
swatch(P63)
names(P63) <- NULL
#Make barplots with different seeds
seed <- 1 #find in dir.create(paste0(dir_out, "/seedopt/"))
set.seed(seed)  
p <- ggplot(df, aes(x = sample, y = row_count, fill = AnnotationRefined)) +
  geom_bar(stat = 'identity', position = 'fill') +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_manual(values = sample(P63)) +
  xlab('Sample ID') +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(colour = "lightgray", size = 0.5)
  ) +
  facet_grid(AnnotationMain ~ ., scales = "free_y")
ggsave(paste0(dir_out, "/seedopt/Barplot_sample_AnnotationRefined_main5_", seed, ".pdf"),
       plot = p,
       width = 15, height = 6)

##Consider leaving out
a1 <- ggplot(df %>% select(sample, EndometriosisStatus) %>% distinct (), aes(x = sample, y = EndometriosisStatus, fill = EndometriosisStatus)) +
  geom_bar(stat = 'identity', position = 'fill') +
  xlab('Sample ID') +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(colour = "lightgray", size = 0.5)
  ) 
a2 <- ggplot(df %>% select(sample, MenstrualCyclePhase) %>% distinct (), aes(x = sample, y = MenstrualCyclePhase, fill = MenstrualCyclePhase)) +
  geom_bar(stat = 'identity', position = 'fill') +
  xlab('Sample ID') +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1),
    panel.grid.minor = element_blank(),
    panel.grid.major.y = element_line(colour = "lightgray", size = 0.5)
  )
a1 + a2 + plot_layout(nrow = 2)
ggsave(paste0(dir_out, "Barplot_sample_AnnotationRefined_main5_annotation.pdf"),
  plot = last_plot(),
  width = 15, height = 6)

@@

Figure 2 e
```{r plotting of cell types per sample}
##########################################################################
#BoxPlots sample, AnnotationMain, EndometriosisGrade, MenstrualCyclePhase_main
df2 <- meta%>%
  filter(Minor.exclusion.criteria.logical == "FALSE") %>%
  group_by(sample, AnnotationMain, EndometriosisGrade, MenstrualCyclePhase_main, EndometriosisStatus) %>% #no sample, AnnotationMain combinations are dropped
  summarise(row_count = n()) %>%
  ungroup() %>%
  filter(!is.na(AnnotationMain)) %>%
  group_by(sample) %>%
  mutate(AnnotationMain_perc = row_count / sum(row_count) * 100) %>%
  ungroup()

#BoxPlots percentage, EndometriosisStatus #for publication
ggplot(df2, aes(x = MenstrualCyclePhase_main, y = AnnotationMain_perc, fill = EndometriosisStatus)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(colour = EndometriosisStatus),
             colour="black",pch=21, 
             position = position_jitterdodge()) +
  RotatedAxis() +
  xlab('Sample ID') +
  facet_wrap(~AnnotationMain,scales = "free", ncol = 6)
ggsave(paste0(dir_out, "Boxplot_MenstrualCyclePhase_main_main5_sample_EndometriosisStatus_perc_publication.pdf"),
  plot = last_plot(),
  width = 12, height = 4.4)

###Significance testing CTL vs Endo
# Perform t-test CTL vs Endo for each cell type grouped by AnnotationRefined, MenstrualCyclePhase_main
df2_ttest_EndometriosisStatus <- df2 %>%
  group_by(AnnotationMain, MenstrualCyclePhase_main) %>%
  summarise(p.value = t.test(AnnotationMain_perc ~ EndometriosisStatus)$p.value) %>%
  ungroup()
#epithelial, proliferative just not significant (p.value = 0.0898)

###Significance testing prolif vs periov vs sec, CTL + Endo
## Perform t-test Prolif vs Peri for each cell type
df2_ttest_ProlifPeri <- df2 %>%
  dplyr::filter(MenstrualCyclePhase_main %in% c("proliferative","periovulatory")) %>%
  group_by(AnnotationMain) %>%
  summarise(p.value = t.test(AnnotationMain_perc ~ MenstrualCyclePhase_main)$p.value) %>%
  ungroup()
#none significant
## Perform t-test Prolif vs Sec for each cell type
df2_ttest_ProlifSec <- df2 %>%
  dplyr::filter(MenstrualCyclePhase_main %in% c("proliferative","secretory")) %>%
  group_by(AnnotationMain) %>%
  summarise(p.value = t.test(AnnotationMain_perc ~ MenstrualCyclePhase_main)$p.value) %>%
  ungroup()
#epi (0.00977) and str (0.0255) significant

## Perform t-test periovulatory vs sec for each cell type
df2_ttest_PeriSec <- df2 %>%
  dplyr::filter(MenstrualCyclePhase_main %in% c("periovulatory","secretory")) %>%
  group_by(AnnotationMain) %>%
  summarise(p.value = t.test(AnnotationMain_perc ~ MenstrualCyclePhase_main)$p.value) %>%
  ungroup()
#epi (0.00794) and str (0.0137) significant
```

Supplementary Data Figure 4b, delete?
```{r plotting of cell types per sample}
##########################################################################################
##Plot differences for DEG analysis samples
#Extract sample names of samples used for DEG analysis
#DEGsamples <- unique(read.csv("/home/common/data/output/projects/ENDO/E044/A030/major_celltypes_Tan/celltype_freq_sample.csv")$Var2)

DEGsamples <- SupplementaryTable1 %>%
  filter(DEG.analysis..Figure.3a.b. == TRUE) %>%
  pull(Patient.sample.ID)

#Make df 
#Introduce 0 for the ones with 0 count!
df3 <- meta%>%
  group_by(sample, AnnotationRefined) %>%
  summarise(AnnotationRefined_counts = n()) %>%
  ungroup() %>%
  complete(sample, AnnotationRefined, fill = list(AnnotationRefined_counts = 0)) %>%
  filter(!is.na(AnnotationRefined)) %>%
  group_by(sample) %>%
  mutate(AnnotationRefined_perc = AnnotationRefined_counts / sum(AnnotationRefined_counts) * 100) %>%
  ungroup()  %>% #60 samples * 63 cell types = 3780 rows, correct
  filter(sample %in% DEGsamples) #23 samples * 63 cell types = 1449 rows, correct
df3 <- left_join(df3, dplyr::distinct(meta[,c("sample","EndometriosisStatus")]), by = "sample")

# Perform t-test for each cell type
df3_ttest <- df3 %>%
  group_by(AnnotationRefined) %>%
  summarise(p.value = t.test(AnnotationRefined_perc ~ EndometriosisStatus)$p.value) %>%
  ungroup()
#No significant difference between the EndometriosisStatus in any cell type

###Plot
#BoxPlots percentage, EndometriosisStatus, percentage
ggplot(df3, aes(x = AnnotationRefined, y = AnnotationRefined_perc, fill = EndometriosisStatus)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(colour = EndometriosisStatus),
             colour="black",pch=21, 
             position = position_jitterdodge(jitter.width = 0.2)) +
  facet_wrap(~AnnotationRefined, scales = "free", ncol = 7)
ggsave(paste0(dir_out, "Boxplot_DEGsamples_SymphonyRefinedFinal_EndometriosisStatus_perc.pdf"),
  plot = last_plot(),
  width = 17, height = 12)
```

Supplementary Data Figure 4b
```{r plotting of cell types per sample}
##########################################################################################
##Plot differences for samples without minor exclusion criteria

#Make df 
#Introduce 0 for the ones with 0 count!
df4 <- meta%>%
  filter(Minor.exclusion.criteria.logical == "FALSE") %>%
  group_by(sample, AnnotationRefined) %>%
  summarise(AnnotationRefined_counts = n()) %>%
  ungroup() %>%
  complete(sample, AnnotationRefined, fill = list(AnnotationRefined_counts = 0)) %>%
  filter(!is.na(AnnotationRefined)) %>%
  group_by(sample) %>%
  mutate(AnnotationRefined_perc = AnnotationRefined_counts / sum(AnnotationRefined_counts) * 100) %>%
  ungroup() %>% #60 samples * 63 cell types = 3780 rows, correct
  filter(!sample %in% c("sample81", "sample87", "sample108", "sample115", "sample119", "sample121"))  #54 samples * 63 cell types = 3402 rows, correct
df4 <- left_join(df4, dplyr::distinct(meta[,c("sample","EndometriosisStatus","MenstrualCyclePhase_main")]), by = "sample")
write.csv(df4, paste0(dir_out,"df4_AnnotationRefined_frequency.csv"))

# Perform t-test CTL vs Endo for each cell type grouped by AnnotationRefined, MenstrualCyclePhase_main
df4_ttest_EndometriosisStatus <- df4 %>%
  group_by(AnnotationRefined, MenstrualCyclePhase_main) %>%
  summarise(p.value = t.test(AnnotationRefined_perc ~ EndometriosisStatus)$p.value) %>%
  ungroup()
#Prv_VSMC secretory, secretory is the only significant one (pval 0.009075467)!!
df4_ttest_EndometriosisStatus$p.value_adjBH <- p.adjust(df4_ttest_EndometriosisStatus$p.value, method = "BH", n = length(df4_ttest_EndometriosisStatus$p.value))
#After correction for multiple testing Prv_VSMC secretory is not significant anymore

df4_ttest_EndometriosisStatus_secretory <- df4_ttest_EndometriosisStatus %>%
                      filter(MenstrualCyclePhase_main == "secretory")
df4_ttest_EndometriosisStatus_secretory$p.value_adjBH <- p.adjust(df4_ttest_EndometriosisStatus_secretory$p.value, method = "BH", n = length(df4_ttest_EndometriosisStatus_secretory$p.value))
#Even when subsetting secretory phase, after correction for multiple testing Prv_VSMC secretory is not significant anymore

## Perform t-test Prolif vs Peri for each cell type
df4_ttest_ProlifPeri <- df4 %>%
  dplyr::filter(MenstrualCyclePhase_main %in% c("proliferative","periovulatory")) %>%
  group_by(AnnotationRefined) %>%
  summarise(p.value = t.test(AnnotationRefined_perc ~ MenstrualCyclePhase_main)$p.value) %>%
  ungroup()
#Few significant
## Perform t-test Prolif vs Sec for each cell type
df4_ttest_ProlifSec <- df4 %>%
  dplyr::filter(MenstrualCyclePhase_main %in% c("proliferative","secretory")) %>%
  group_by(AnnotationRefined) %>%
  summarise(p.value = t.test(AnnotationRefined_perc ~ MenstrualCyclePhase_main)$p.value) %>%
  ungroup()
write.csv(df4_ttest_ProlifSec, paste0(dir_out,"df4_ttest_ProlifSec.csv"))
df4_ttest_ProlifSec <- read.csv(paste0(dir_out,"df4_ttest_ProlifSec.csv"))
df4_ttest_ProlifSec$p.value_adjBH <- p.adjust(df4_ttest_ProlifSec$p.value, method = "BH", n = length(df4_ttest_ProlifSec$p.value))
write.csv(df4_ttest_ProlifSec, paste0(dir_out,"df4_ttest_ProlifSec_adjpval.csv"))

#About half are significant, show in paper supplement
## Perform t-test periovulatory vs sec for each cell type
df4_ttest_PeriSec <- df4 %>%
  dplyr::filter(MenstrualCyclePhase_main %in% c("periovulatory","secretory")) %>%
  group_by(AnnotationRefined) %>%
  summarise(p.value = t.test(AnnotationRefined_perc ~ MenstrualCyclePhase_main)$p.value) %>%
  ungroup()
#Less than half are significant, don't show in paper supplement

#BoxPlots percentage, EndometriosisStatus, percentage ###for paper Extended Data Fig. S4
ggplot(df4, aes(x = MenstrualCyclePhase_main, y = AnnotationRefined_perc, fill = EndometriosisStatus)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(colour = EndometriosisStatus),
             colour="black",pch=21, 
             position = position_jitterdodge(jitter.width = 0.2)) +
  facet_wrap(~AnnotationRefined, scales = "free", ncol = 7) +
  theme_bw() +
  expand_limits(y = 0)

ggsave(paste0(dir_out, "Boxplot_DEGsamples_SymphonyRefinedFinal_EndometriosisStatus_perc.pdf"),
  plot = last_plot(),
  width = 20, height = 12)
```

Figure 2f
```{r plotting of cell types per sample}
#Early and Mid Secretory Prv_VSMCs 
df4 <- meta%>%
  filter(Minor.exclusion.criteria.logical == "FALSE") %>%
  group_by(sample, AnnotationRefined) %>%
  summarise(AnnotationRefined_counts = n()) %>%
  ungroup() %>%
  complete(sample, AnnotationRefined, fill = list(AnnotationRefined_counts = 0)) %>%
  filter(!is.na(AnnotationRefined)) %>%
  group_by(sample) %>%
  mutate(AnnotationRefined_perc = AnnotationRefined_counts / sum(AnnotationRefined_counts) * 100) %>%
  ungroup() %>% #60 samples * 63 cell types = 3780 rows, correct
  filter(!sample %in% c("sample81", "sample87", "sample108", "sample115", "sample119", "sample121"))  #54 samples * 63 cell types = 3402 rows, correct
df4 <- left_join(df4, dplyr::distinct(meta[,c("sample","EndometriosisStatus","MenstrualCyclePhase_main","MenstrualCyclePhase")]), by = "sample")

secEarylMid_Prv_VSMC_MenstrualCyclePhase_main <- df4 %>%
  filter(AnnotationRefined %in% c("Prv_VSMC secretory", "Prv_VSMC secretory_late")) %>% #same result with and without "Prv_VSMC secretory_late"
  dplyr::filter(MenstrualCyclePhase %in% c("proliferative","periovulatory","secretory_mid","secretory_early")) %>%
  group_by(sample, EndometriosisStatus, MenstrualCyclePhase_main) %>%
  summarize(
    total_counts_MenstrualCyclePhase_main = sum(AnnotationRefined_counts, na.rm = TRUE),
    total_perc_MenstrualCyclePhase_main = sum(AnnotationRefined_perc, na.rm = TRUE)
  ) %>%
  ungroup()

ggplot(secEarylMid_Prv_VSMC_MenstrualCyclePhase_main, aes(x = MenstrualCyclePhase_main, y = total_perc_MenstrualCyclePhase_main, fill = EndometriosisStatus)) +
  geom_boxplot(outlier.shape = NA) +
  geom_point(aes(colour = EndometriosisStatus),
             colour="black",pch=21, 
             position = position_jitterdodge(jitter.width = 0.2))
ggsave(paste0(dir_out, "Boxplot_sec_Prv_VSMC_NOsecretory_late_total_perc_MenstrualCyclePhase_main_EndometriosisStatus.pdf"),
  plot = last_plot(),
  width = 17, height = 12)

secEarylMid_Prv_VSMC_MenstrualCyclePhase_main_pval <- secEarylMid_Prv_VSMC_MenstrualCyclePhase_main %>%
  dplyr::filter(MenstrualCyclePhase_main %in% c("secretory")) %>% 
  summarise(p.value = t.test(total_perc_MenstrualCyclePhase_main ~ EndometriosisStatus)$p.value) %>%
  ungroup()
#significant, p.value 0.0261 -> paper

secEarylMid_Prv_VSMC_MenstrualCyclePhase_main_pval_prolif <- secEarylMid_Prv_VSMC_MenstrualCyclePhase_main %>%
  dplyr::filter(MenstrualCyclePhase_main %in% c("proliferative")) %>% 
  summarise(p.value = t.test(total_perc_MenstrualCyclePhase_main ~ EndometriosisStatus)$p.value) %>%
  ungroup()
#significant, p.value NA

secEarylMid_Prv_VSMC_MenstrualCyclePhase_main_pval_peri <- secEarylMid_Prv_VSMC_MenstrualCyclePhase_main %>%
  dplyr::filter(MenstrualCyclePhase_main %in% c("periovulatory")) %>% 
  summarise(p.value = t.test(total_perc_MenstrualCyclePhase_main ~ EndometriosisStatus)$p.value) %>%
  ungroup()
#significant, p.value 0.7639853

p.adjust(c(0.0261), method = "BH", n = length(3)) #Value does not change
p.adjust(c(0.7639853), method = "BH", n = length(3)) #Value does not change

#Calculate fold change ENDO vs nonENDO secretory
# Assuming secEarylMid_Prv_VSMC_MenstrualCyclePhase_main is your dataframe

library(dplyr)

# Filter the dataframe for 'secretory' in MenstrualCyclePhase_main and then group by EndometriosisStatus
mean_total_perc <- secEarylMid_Prv_VSMC_MenstrualCyclePhase_main %>%
  filter(MenstrualCyclePhase_main == "secretory") %>%
  group_by(EndometriosisStatus) %>%
  summarize(mean_total_perc = mean(total_perc_MenstrualCyclePhase_main))
```

Figure 1d
```{r plotting of cell types per sample}
p5 <- ggplot(df5, aes(x = AnnotationMain, y = row_count)) +
  geom_col() +
  #theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  coord_flip()
p5
ggsave(paste0(dir_out, "Barplot_EndometriosisGrade_main5.pdf"),
  plot = last_plot(),
  width = 15, height = 6)

#main5 boxplot 
df6 <- meta%>%
  group_by(sample, AnnotationMain, EndometriosisGrade) %>%
  summarise(row_count = n()) %>%
  ungroup()
p6 <- ggplot(df6, aes(x = AnnotationMain, y = row_count)) +
  geom_boxplot() +
  #geom_jitter(position = position_jitter(width = 0.2, height = 0), size = 0.5) +
  scale_y_log10() +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
p6
ggsave(paste0(dir_out, "Boxplot_EndometriosisGrade_main5_colEndometriosisGrade.pdf"),
         plot = last_plot(),
         width = 15, height = 6)

# bar plot of the number of AnnotationRefined cells in each sample
df7 <- meta%>%
  filter(!is.na(AnnotationMain)) %>%
  group_by(AnnotationMain, EndometriosisGrade) %>%
  summarise(row_count = n()) %>%
  ungroup() 
ggplot(df7, aes(x = AnnotationMain, y = row_count, fill = EndometriosisGrade)) +
  geom_bar(stat = 'identity', position = 'fill') +
  scale_y_continuous(expand = c(0, 0)) +
  coord_flip()
ggsave(paste0(dir_out, "Barplot_AnnotationMain_EndometriosisGrade.pdf"),
  plot = last_plot(),
  width = 6, height = 4)

@@

UMAPs

epithelial UMAP
2b,2c,3b,3e,

atlas UMAP
S3f, S7a

Figure 1c
```{r}
data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033.rds")
#data <- readRDS("/home/common/data/output/projects/ENDO/E044/A033/dataA003A033_downsampled.rds")
A003A033 <- readRDS("/home/common/data/output/projects/ENDO/E044/A074/dataA003A033_metadata_ptime_snn_Final.rds")
table(A003A033$AnnotationRefined)
data <- AddMetaData(data,A003A033)
E031A004meta <- read.csv("/home/common/data/output/projects/ENDO/E044/A033/E031A004_proliferative_SCT_percent.mt_sample_integration_batch_hvg_annotated_meta.csv")
E031A004meta$cell_id <- paste(E031A004meta$sample, E031A004meta$cell.names, sep = "_")
rownames(E031A004meta) <- E031A004meta$cell_id 
data <- AddMetaData(data, E031A004meta[, "Annotation", drop = FALSE])

>>>>

@@

ln -s /home/common/data/output/projects/ENDO/E044/A033/EndoAtlas_GEO_meta_complete_downsampled.rds EndoAtlas.rds
EndoAtlas_meta <- readRDS(paste0(dir_data,"EndoAtlas_meta.rds"))
EndoAtlas <- readRDS(paste0(dir_data,"EndoAtlas.rds"))
EndoAtlas <- AddMetaData(EndoAtlas, EndoAtlas_meta)
saveRDS(EndoAtlas,paste0(dir_data,"EndoAtlas.rds") )

Add additional data to the Seurat metadata and reformat
```{r plotting of cell types per sample}
#Set directory links and create directories
dir_ENDO <- "/home/duempelmann/analysis-projects/ENDO/E044/A085/endometriosis_endometrium_scRNA_atlas/"

dir_data <- paste0(dir_ENDO,"_Data/")
dir_out <- paste0(dir_ENDO,"Figures_out/")
if (!dir.exists(dir_out)) {
  dir.create(dir_out, recursive = TRUE)
}

#Load data
EndoAtlas <- readRDS(paste0(dir_data, "EndoAtlas.rds"))
EndoAtlas_meta <- EndoAtlas@meta.data

#Preserve cell names
EndoAtlas_meta$cell_names <- rownames(EndoAtlas_meta)

##Modify menstrual cycle information
# Create column MenstrualCyclePhase_main based on MenstrualCyclePhase by merging the secretory sub-phases from 
EndoAtlas_meta$MenstrualCyclePhase_main <- as.character(EndoAtlas_meta$MenstrualCyclePhase)
EndoAtlas_meta$MenstrualCyclePhase_main[EndoAtlas_meta$MenstrualCyclePhase_main %in% c("secretory_early", "secretory_mid", "secretory_late")] <- "secretory"
#Factor samples by MenstrualCyclePhase_main for plotting
EndoAtlas_meta$MenstrualCyclePhase_main <- factor(EndoAtlas_meta$MenstrualCyclePhase_main, levels = c("proliferative","periovulatory","secretory"))

##Add additional metadata from manuscript Supplementary Table 1
# Read the tab-delimited text file
SupplementaryTable1 <- read.delim(paste0(dir_data, "SupplementaryTable1.txt"), header = TRUE, stringsAsFactors = FALSE)
# Perform a left join of the `meta` dataframe with `SupplementaryTable1`
EndoAtlas_meta <- dplyr::left_join(
  EndoAtlas_meta,
  SupplementaryTable1 %>% 
    dplyr::select(
      Patient.sample.ID,
      epithelial.pseudotime,
      sequencing.batch,
      X10x.capturing.batch,
      Minor.exclusion.criteria.logical,
      Menstrual.cycle.phase.marker.analysis,
      Cell.type.frequency.analysis..Figure.2e.f.,
      DEG.analysis..Figure.3a.b.,
      Ligand.receptor.analysis..Figure.3c.h.
    ),
  by = c("sample" = "Patient.sample.ID")
)

##Determine order of samples by MenstrualCyclePhase and epithelial.pseudotime
MenstrualCyclePhase_sampleOrder <- EndoAtlas_meta%>%
  select(sample, MenstrualCyclePhase,epithelial.pseudotime) %>%
  distinct() %>%
  arrange(MenstrualCyclePhase, epithelial.pseudotime) %>%
  mutate(MenstrualCyclePhase_sampleOrder = row_number()) %>%
  ungroup()
#Factor samples by MenstrualCyclePhase and epithelial.pseudotime for plotting
EndoAtlas_meta$sample <- factor(EndoAtlas_meta$sample, levels = c(MenstrualCyclePhase_sampleOrder$sample))

#Add cell names as row names
rownames(EndoAtlas_meta) <- EndoAtlas_meta$cell_names

#Save EndoAtlas_meta as .rds
saveRDS(EndoAtlas_meta, paste0(dir_data, "EndoAtlas_meta.rds"))

#EndoAtlas_meta <- readRDS(paste0(dir_data, "EndoAtlas_meta.rds"))

@@

Save epithelial object and upload to GEO.
```{r}
#Load monocl object
data_monocl <- readRDS("/home/common/data/output/projects/ENDO/E044/A062/trajectory_day5-6-7_greenCircle/trajectory_timepoint0.rds")
listData <- (data_monocl@colData@listData)
listData <- listData["cell_ident"]

EndoAtlas_meta <- readRDS(paste0(dir_data, "EndoAtlas_meta.rds"))
colnames(EndoAtlas_meta)

########################
# Extract cell.ident from listData
cell_ids <- listData[["cell_ident"]]

# Ensure cell_ids is not empty
if (length(cell_ids) == 0) {
  stop("cell.ident in listData is empty. Ensure it is populated before proceeding.")
}

# Filter and order EndoAtlas_meta by cell_ids
filtered_meta <- EndoAtlas_meta[EndoAtlas_meta$cell_names %in% cell_ids, ]
ordered_meta <- filtered_meta[match(cell_ids, filtered_meta$cell_names), ]

# Verify alignment
if (!all(ordered_meta$cell_names == cell_ids)) {
  stop("Ordering of filtered EndoAtlas_meta does not match cell_ids.")
}

# Add columns from ordered_meta to listData
columns_to_add <- c("sample", "EndometriosisStatus", "EndometriosisGrade", "CycleDay", 
                    "ProgesteroneSerum", "MenstrualCyclePhase", "AnnotationMain", 
                    "AnnotationRefined", "cell_names", "MenstrualCyclePhase_main")

for (col in columns_to_add) {
  if (col %in% colnames(ordered_meta)) {
    listData[[col]] <- ordered_meta[[col]]
  } else {
    warning(paste("Column", col, "not found in EndoAtlas_meta. Skipping."))
  }
}

# Update the main object
data_monocl@colData@listData <- listData

data_monocl@principal_graph_aux@listData[["UMAP"]][["pseudotime"]]

saveRDS(data_monocl, paste0(dir_data, "epithelial_cells_monocle3.rds"))
```

###########################################
Main Figures
###########################################

Figure 1c
UMAP of the entire endometrium single-cell atlas from women with and without endometriosis, colored by cell types (n=63).
```{r}
# Define order of cell types
order <- EndoAtlas_meta %>%
  select(AnnotationMain, AnnotationRefined) %>%
  filter(!is.na(AnnotationMain)) %>%
  distinct() %>%
  arrange(AnnotationRefined) %>%
  mutate(AnnotationMain = factor(AnnotationMain, levels = c( "endothelial","stromal","lymphocyte","epithelial","myeloid"))) %>%
  arrange(AnnotationMain)
EndoAtlas$AnnotationRefined <- factor(x = EndoAtlas$AnnotationRefined, levels = order$AnnotationRefined)
# DimPlot
Fig1c <- DimPlot(EndoAtlas, reduction = "umap", group.by = "AnnotationRefined",  raster = FALSE)
Fig1c
# Save DimPlot
ggsave(filename = paste0(dir_out, 'Fig1c.pdf'),
       plot = Fig1c,
       width =20, height=9)

@@

Figure 1d
Main cell type frequencies per patient (n=60, box plot, left) and entire dataset (bar plot, middle), with mesenchymal cells being the most abundant cell type. The overall cell number distribution between non-ENDO, mild-ENDO and severe-ENDO is balanced (bar plot, right).
```{r plotting of cell types per sample}
## Fig1d_1 Cells per patient
# Boxplot
Fig1d_1 <- EndoAtlas_meta %>%
  group_by(AnnotationMain) %>%
  summarise(row_count = n()) %>% 
  ggplot(aes(x = AnnotationMain, y = row_count)) +
  geom_col() +
  coord_flip()
Fig1d_1
# Save Boxplot
ggsave(paste0(dir_out, "Fig1d_1.pdf"),
  plot = last_plot(),
  width = 15, height = 6)

## Fig1d_2 Cells per main cell type
# Barplot
Fig1d_2 <- EndoAtlas_meta %>%
  group_by(sample, AnnotationMain) %>%
  summarise(row_count = n()) %>%
  ggplot(aes(x = AnnotationMain, y = row_count)) +
  geom_boxplot() +
  scale_y_log10() +
  coord_flip() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
Fig1d_2
# Save Barplot
ggsave(paste0(dir_out, "Fig1d_2.pdf"),
         plot = last_plot(),
         width = 15, height = 6)

##Fig1d_3 cell number distribution between non-ENDO, mild-ENDO and severe-ENDO
# Barplot
Fig1d_3 <- EndoAtlas_meta%>%
  filter(!is.na(AnnotationMain)) %>%
  group_by(AnnotationMain, EndometriosisGrade) %>%
  summarise(row_count = n()) %>%
  ggplot(aes(x = AnnotationMain, y = row_count, fill = EndometriosisGrade)) +
  geom_bar(stat = 'identity', position = 'fill') +
  scale_y_continuous(expand = c(0, 0)) +
  coord_flip()
Fig1d_3
# Save Barplot
ggsave(paste0(dir_out, "Fig1d_3.pdf"),
  plot = last_plot(),
  width = 6, height = 4)

@@

Figure 1e
Eigencor plot showing the effect of different co-variants of the atlas dataset on principal components (PCs) with Pearson correlation values and multiple-testing adjustment with the Benjamini-Hochberg procedure.

CODE running, need to extend the PCA analysis folder
dir_out <- "/home/common/data/output/projects/ENDO/E044/A072/allcells_EndoAtlas_SCTassay/"
```{r}
## Figure 1e
# Load data
p <- readRDS(paste0(dir_ENDO,"_Data/03_PCA_analysis/p_EndoAtlas.rds"))
# Plot and save eigencorplot
pdf(paste0(dir_out,"eigencorplot_final_BH.pdf"), width = 10, height = 5)
eigencorplot(p,
    metavars = c("sample","EndometriosisStatus","EndometriosisGrade","CycleDay","ProgesteroneSerum","MenstrualCyclePhase","MenstrualCyclePhase_main","sequencing.batch", "X10x.capturing.batch"),
    col = c('darkorange3','darkorange', 'white','cadetblue2','cadetblue'),
    corMultipleTestCorrection = "BH"
)
dev.off()
```
